Modeling costs and benefits of genomic wastewater surveillance

### 2023-05-31 paper/manuscript.qmd
implementing NPIs with a lagged outcome
writing up travel and stochastic implementation equations

$$ 
\begin{align}
\dot{\mathbf{S}} &=\ - \boldsymbol{\lambda}_t \odot \mathbf{S} + p_S \mathbf{m}_S \\
\dot{\mathbf{E}} &=\ \boldsymbol{\lambda}_t \odot \mathbf{S} - \sigma \mathbf{E} + p_E \mathbf{m}_E \\
\dot{\mathbf{P}} &=\ \sigma \mathbf{E} - \delta \mathbf{P} + p_P \mathbf{m}_P \\
\dot{\mathbf{I}} &=\ \delta \mathbf{P} - \gamma \mathbf{I} + p_I \mathbf{m}_I \\
\dot{\mathbf{R}} &=\ \gamma \mathbf{I} + p_R \mathbf{m}_R \\
\end{align}
$$ {#eq-ode1}

where $\dot{\mathbf{S}}, \dot{\mathbf{E}}, \dot{\mathbf{P}}, \dot{\mathbf{I}}, \dot{\mathbf{R}}$ represent the time derivatives of each disease status, greek letters represent progression rates and $p_x\mathbf{m}_x$ terms represent migration flows $\mathbf{m}_x$ modulated by propensity to travel $p_x$ at disease state $x \in \{S, E, P, I, R\}$.

@@

<!--
### Discrete-time stochastic implementation

The model described above is implemented as a discrete-time stochastic model using the following approach. We simulate outflows for each disease state vector $\mathbf{X}$ using a stochastic  $\mathbf{X_t} = \mathbf{X_{t-1}} \sim Binomial(\mathbf{X}, 1-exp(\mathbf{p}))$ hence.

@@

$$ 
\boldsymbol{\beta}_t = (1-\mathbf{l}_t\tau) \boldsymbol{\beta}_b
$$ {#eq-betat}

@@

$$ 
\mathbf{l}_t^* = \min{(cI, l_{max})}
$$ {#eq-targetnpi}

This formulation approximates phased mitigation plans adopted during the COVID-19 pandemic, where policymakers introduce or remove interventions depending on disease prevalence or other severity indicators. While public health decision-makers strive to use up-to-date information to introduce and remove NPIs, they do not have perfect and instantaneous information on disease prevalence; hence actual policy behaviors lag epidemiological indicators. In continuous time, we represent this process with a stock vector $\mathbf{L}$ representing the NPI level in effect

@@

where $a$ represents the adjustment time for NPIs (i.e., one or two weeks) using conventional measures.

>>>>

$$ 
\begin{align}
\dot{\mathbf{S}} &=\ - \boldsymbol{\lambda}_t \odot \mathbf{S} + \mathbf{m}_S \\
\dot{\mathbf{E}} &=\ \boldsymbol{\lambda}_t \odot \mathbf{S} - \sigma \mathbf{E} + \mathbf{m}_E \\
\dot{\mathbf{P}} &=\ \sigma \mathbf{E} - \delta \mathbf{P} + \mathbf{m}_P \\
\dot{\mathbf{I}} &=\ \delta \mathbf{P} - \gamma \mathbf{I} + \mathbf{m}_I \\
\dot{\mathbf{R}} &=\ \gamma \mathbf{I} + \mathbf{m}_R \\
\end{align}
$$ {#eq-ode1}

where $\dot{\mathbf{S}}, \dot{\mathbf{E}}, \dot{\mathbf{P}}, \dot{\mathbf{I}}, \dot{\mathbf{R}}$ represent the time derivatives of each disease status, greek letters represent progression rates and $\mathbf{m}_x$ terms represent net travel flows $\mathbf{m}_x$ at disease state $x \in \{S, E, P, I, R\}$.

@@

### Discrete-time stochastic implementation

We implement the model as a discrete-time stochastic model using the following approach. We simulate outflows for each disease state vector $\mathbf{X}$ using drawing from a binomial distributions $\mathbf{out_{X_t}} \sim Binomial(\mathbf{X}, 1-exp(\mathbf{p}))$. We then carry over outflows from one disease state to the next. Hence, our disease state equations can be re-written as

$$ 
\begin{align}
\mathbf{S_t} &\sim \mathbf{S_{t-1}} 
                    -B(\mathbf{S_{t-1}},1-\exp(-\boldsymbol{\lambda}_{t-1}/\mathbf{N}))
                      + \mathbf{m_S} \\
\mathbf{E_t} &\sim \mathbf{E_{t-1}} 
                    +B(\mathbf{S_{t-1}},1-\exp(-\boldsymbol{\lambda}_{t-1}/\mathbf{N}))
                    -B(E_{t-1},1-\exp(-\sigma)) 
                    +\mathbf{m_E}\\
\mathbf{P_t} &\sim \mathbf{P_{t-1}}
                    +B(E_{t-1},1-\exp(-\sigma))
                    -B(P_{t-1},1-\exp(-\delta)) +
                    \mathbf{m_P}\\
\mathbf{I_t} &\sim \mathbf{I_{t-1}}
                    +B(P_{t-1},1-\exp(-\delta))
                    -B(I_{t-1},1-\exp(-\gamma)) +
                    \mathbf{m_I}\\
\mathbf{R_t} &\sim \mathbf{R_{t-1}}
                    -B(I_{t-1},1-\exp(-\gamma)) +
                    \mathbf{m_R}
\end{align}
$$ {#eq-stochastic}

where $B(\mathbf{n},p)$ results in a binomial draw from population vector $n$ with probability $p$. The model simulates outflows and inflows such that the outflows from one disease state are the same as the inflows to a different state.

### Travel

Stochastic travel flows are modeled as follows. Mobility matrix $\mathbf{A}$ is an $n \times n$ matrix where $a_{ij}$ represent daily travel rates from jurisdiction $i$ to $j$ and $a_{i=j}$ is the probability of no travel. Travel between jurisdictions is modeled using multinomial draws for each source jurisdiction $i$:

$$
\mathbf{X'_{td}} \sim \sum_{j}{Multinom(\mathbf{X_{t-1}}, A_i )} \ \forall \ d \in \{S, E, P, I, R\}
$$ {#eq-travel}
where $\mathbf{X'_{td}}$ are disease state vectors after travel. Once travel is computed in the model, we back out net travel flows as $\mathbf{m_d} = \mathbf{X'_{td}} - \mathbf{X_{td}}$ and use those in @eq-stochastic. This formulation implies that travel is simulated before disease transitions *within* each time step.

@@

$$ 
\boldsymbol{\beta}_t = (1-\mathbf{L}_t\tau) \boldsymbol{\beta}_b
$$ {#eq-betat}

@@

$$ 
\mathbf{L}_t^* = \min{(cI_{t-\Delta}, l_{max})}
$$ {#eq-targetnpi}
where $\Delta$ is the delay imposed by the surveillance method used to monitor the disease. This formulation approximates phased mitigation plans adopted during the COVID-19 pandemic, where policymakers introduce or remove interventions depending on disease prevalence or other severity indicators. While public health decision-makers strive to use up-to-date information to introduce and remove NPIs, they do not have perfect and instantaneous information on disease prevalence; hence actual policy behaviors lag epidemiological indicators. In continuous time, we represent this process with a stock vector $\mathbf{L}$ representing the NPI level in effect

@@

where $a$ represents the adjustment time for NPIs for decision making reasons, whereas $\Delta$ represents the lag imposed by surveillance methods.
